Numerical modelling of the Cordilleran ice sheet

### 2014-07-17 kappa.tex
Update author contrib, and correct broken refs.

\begin{description}
  \item[\CCLI] \bibentry{Seguinot.etal.2014}
  \item[\PSDV] \bibentry{Seguinot.2013}
  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}
  \item[\CCYC] \bibentry{Seguinot.etal.inpreparation}
\end{description}

@@

\small{\begin{enumerate}
  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
simulations and wrote the manuscript; C.~Khroulev implemented necessary code
changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
some of the forcing data. All co-authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model,
ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
reviews.
  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
data; and we wrote the manuscript together.
  \item[IV] I implemented necessary code changes in PISM, ran the simulations
and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
contributed to improve the text.
\end{enumerate}}

@@

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}).
During the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
ranges of the Western Cordillera, which forms our region of focus in the
present thesis.

@@

The North American Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
It was formed during multiple accretion phases associated with the closure of
several back-arc oceanic trenches, over a 200\,Ma-long geological history
\citep{Sigloch.Mihalynuk.2013}.

Most of the dominant reliefs are found along two major ribbons that run
parallel to each other along the general orientation of the Cordillera
(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
orogenic belt \mref. Major glaciated reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The Cordillera comports several intermontane basins, which influenced the
patterns of ice flow. The two major depressions are the Fraser Plateau in
south-central British Columbia, and the Liard Lowland in northern British
Columbia. Between these two basins is found a mildly elevated, but extensive
and deeply incised mountain group, known as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them
\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
modelling, this topography implies complex patterns of flow and stress
within the ice and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, \emph{mean annual temperatures} vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However,
\emph{seasonality},
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Rates of \emph{precipitation} are highest on the western slopes of the Pacific
Coast
ranges and decline quickly on the eastern side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the \emph{precipitation peak}
through
the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation,
consequently falls there
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

The first presence of an ice sheet in the Cordilleran mountains is indicated by
glaciofluvial deposits dated at $2.64^{+0.20}/_{-0.18}$\,Ma
\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
been glaciated many times, with the first glaciations occurring even before the
onset of the 100-ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
the last few glaciations can be reconstructed from the geologic record, whereas
much of the older evidence has been overridden by subsequent glacial
re-advances and other geomorphological processes.

In the \emph{northern sector} of the ice sheet, at least four glaciations have
been identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid glaciations,
pre-date the last glacial cycle, and thus are beyond the scope of this study.
The two younger glaciations, though, occurred both during the last glacial
cycle. First, the Gladstone glaciation reached its maximum between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during a period of high global ice volume on Earth, known as the Marine Isotope
Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
found in the southern part of the ice-covered region, making it difficult to
seize its magnitude. Second, the McConnell glaciation is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which corresponds to the last global ice volume maximum (MIS~2). In contrary to
the Gladstone glaciation, the McConnell glaciation has been correlated with a
contemporaneous advance of the southern ice sheet margin, known as the Fraser
glaciation, with a maximum extent framed between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
This is the period where most geological observations point at.

It is generally thought that the Cordilleran ice sheet was short-lived
    \citep[e.g.]{Clague.etal.1980, Clague.1985, Cosma.etal.2008}.
The ice sheet originated from the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
During the last glacial maximum, ice covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}).

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of different piedmont
lobes that have formed at the outlet of the main valleys. The Puget lobe
\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The \emph{western margin} of the ice sheet covered part of the continental
shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent \mref. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula where
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely indicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier \mref.

Finally, the \emph{eastern margin} of the ice sheet is the most unclear.
Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the nature of the interaction between the two ice sheets and the precise
location of this junction is unclear, while little information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, our largest
assumption certainly consists of neglecting buttressing effects from the
Laurentide ice sheet.

The ice sheet is believed to have retreated quickly. Most of the evidence comes
from the Fraser Plateau, where it indicates that some of the highest relief
emerged from the ice prior to the lowlands, possibly resulting in thin,
stagnant ice remnants in the interior
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Much less is known about ice sheet decay in the northern parts of the
ice-covered areas, but glacial landforms from the Liard Lowland suggest a
different mode of \emph{deglaciation}, with active retreat of the ice lobe
towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.

@@

The simulations presented in this thesis use Parallel Ice Sheet Model (PISM),
an open-source, finite-difference, shallow ice-sheet model
\citep{PISM-authors.2014}. The model reads bedrock topography, sea level,
geothermal heat flux and climate forcing as inputs, and computes the thermal
and dynamic state of the ice sheet, and the associated thermal and
deformational response of an idealised lithosphere. The atmospheric and oceanic
envelopes are not part of the model. However, PISM includes boundary models
that provide simple parametrisations of their effect on the ice model at
supposedly well-defined interfaces (\cref{fig:model-interfaces}).

This section does not attempt to provide an extensive description of PISM,
which can be found in the on-line documentation \citep{PISM-authors.2014}.
Instead, it aims to highlight modelling choices made in the context of this
thesis, and give an overview of the physics involved in these choices.

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, pressure, and
water content through an Arrhenius-type law
\citep[Eqs.~63--65]{Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
capped at a threshold of 0.01, above which no measurements are available
\citep{Lliboutry.Duval.1985}.

\todo{I am trusting PISM's manual on that. Check other papers by Duval, or
      the book by Greve (library).}

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
account for thermodynamic effects associated with internal phase changes in the
glacier. This enthalpy ($H$) formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,,
\end{equation}
where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
conductivity. Parameter values for the ice flow model used in this thesis are
given in \cref{tab:iceflowparams}.

\Cref{eqn:incompressibility,eqn:stressbalance,eqn:glenslaw,eqn:enthalpy}
constitute the thermodynamic basis
of our model. However, their resolution in full form implies a computational
demand too high for multi-millennial, continental-scale applications, such as
the numerical modelling of the Cordilleran ice sheet through the last glacial
cycle, thus requiring shallow approximations.

@@

PISM, the ice sheet model that we use, employs two well-documented
approximations of ice flow equations: the shallow ice approximation (SIA)
and the the shallow shelf approximation (SSA). Although both approximations can
be derived from a rigorous scaling approach
    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983,
                   Morland.1987, Weis.etal.1999},
we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

Neglecting atmospheric pressure, \Cref{eqn:siaz} yield
\begin{equation}
    p = \rho g (s-z) \,,
\end{equation}
which can be re-introduced into \cref{eqn:siax,eqn:siay}, thus yielding the
expression of the horizontal shear stresses, sometimes referred to as driving
stresses,
\begin{subequations}
\begin{align}
    \tau_{xz} &= -\rho g (s-z) \pdv{s}{x} \,, \\
    \tau_{yz} &= -\rho g (s-z) \pdv{s}{x} \,.
\end{align}
\end{subequations}

@@

Drainage of water at the base of the ice sheet, although implemented in PISM,
is not included in simulations presented here for the sake of simplicity.
In turn, the amount of water at the bed, $W$, corresponds to the local
accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
above which further melt simply exits the system.

>>>>

\begin{description}
  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
  \item[\PSDV] \bibentry{Seguinot.2013}.
  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.\renote{
    Arjen, you had several comments regarding the style of references
    throughout the kappa (name formatting, italic text). I am using the
    Copernicus style here, but is there another recommendation to follow?
    Maybe Peter knows?}
\end{description}

@@

\footnotesize{\begin{enumerate}
  \item[I] A.~P.~Stroeven initiated the Cordilleran ice sheet modelling
project; I set-up the ice sheet model (PISM) and supercomputing workflow, ran
simulations and wrote the first draft; C.~Khroulev implemented necessary code
changes in the model; I.~Rogozhina contributed to experiment design; A.~P.~Stroeven
helped with interpretation of the results; and Q.~Zhang prepared some of the
forcing data. All co-authors contributed to improve the text.
  \item[II] I analysed the weather data, wrote the positive degree day model
(PyPDD), ran computations and wrote the manuscript. A.~P.~Stroeven contributed
with editorial comments at the pre-submission stage and I.~Rogozhina helped to
assess the reviews.
  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
data and developed a new parameterization of daily temperature standard
deviation. I.~Rogozhina and I wrote the manuscript. A.~P.~Stroeven contributed
with editorial comments at the pre-submission stage.
  \item[IV] I integrated the new parameterization of standard deviation into
PISM, ran the simulations and wrote the first draft; I.~Rogozhina guided
experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
interpretation and comparison of model results against geological evidence. All
co-authors contributed to improve the text.
\end{enumerate}}

@@

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}).
During the Last Glacial Maximum (LGM), three major ice sheets had collided to
form a continuous blanket of ice that spanned from southern Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Ohio
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet}, intermediate in size,
capped the mountain
ranges of the Western Cordillera.

@@

The North American Cordillera forms an elongated but broad mountain system with
an average east-west width of about 1000\,km (\cref{fig:map-cordillera}). Its
topography is a complex jigsaw of mountain ranges and interior basins. It was
formed during multiple accretion phases associated with the closure of several
back-arc oceanic trenches, over a 200 million year (Ma)-long geological history
\citep{Sigloch.Mihalynuk.2013}.

The dominant reliefs occur along two major ribbons that run
parallel to each other along the general orientation of the North American Cordillera
(\cref{photo-alaska-range}). The western ribbon is formed by the Pacific Coast
orogenic belt \mref\ and includes,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref, depression and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the South. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The North American Cordillera hosts several intermontane basins, the presence
of which influenced former patterns of Cordilleran ice sheet flow. The two
major depressions are the Fraser Plateau in south-central British Columbia, and
the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
These two basins are separated by the Skeena Mountains, a mildly elevated, but
extensive and deeply incised mountain group. This is the most dominant upland
area in-between the dominant topographic ribbons that form the North American
Cordillera.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing the palaeoglaciology of the Cordilleran ice sheet.
Because of the irregular arrangement of valleys and ridges that formed the ice
sheet's subglacial landscape, ice moved along tortuous and intertwined
pathways, sometimes following the major topographical troughs, and sometimes
flowing across them \citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of
numerical ice sheet modelling, the presence of steep and irregular subglacial
topography yields complex flow patterns and stress distributions within the ice
sheet and requires high-resolution and thus, large computational grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, \emph{mean annual temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, \emph{temperature seasonality}\renote{
    This is a common term in the meteorological literature, I think. At least
    this is what Qiong told me.},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors \renote{
    I think the word ``polar'' or ``Arctic'' is needed here, see
    \cref{fig:plot-atm}b.}
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

\emph{Annual precipitation rates} are highest on the western slopes of the
Pacific Coast ranges and decline quickly on the eastern side of the mountains
(\cref{fig:photo-fraser-valley}, \cref{fig:plot-atm}c). In terms of ice sheet
mass balance, this contrast is even reinforced through the difference in timing
of the \emph{precipitation peaks} through the year (\cref{fig:plot-atm}d).
Coastal regions receive most precipitation during winter and fall, which is
therefore largely delivered as snow. In contrast, many inland regions
experience dry winters, and most of the annual precipitation, consequently
falls there as rain during summer, further enhancing ice melt. The climatic
setting of the North American Cordillera is further described in \CCLI, where
its effects on the ice sheet model response have been studied in detail.

@@

Evidence from glaciofluvial deposited dated using cosmogenic nuclides at $2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the Cordilleran mountains in the early Pleistocene
\citep{Hidy.etal.2013}.

This indicates that the North American Cordillera has
been glaciated many times since the late Pliocene, with the first of these glaciations occurring even before the mid-Pleistocene
onset of 100 ka glacial cycles (\cref{fig:paleo-timeseries}). However, only fragments of
the last few glaciations can be reconstructed from the geologic record, implying that
much of the older evidence has been erased by subsequent glacial
re-advances and perhaps even by other, non-glacial geomorphological processes.

In the \emph{northern sector} of the Cordilleran ice sheet domain, at least four glaciations have
been identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid  and pre-Reid glaciations,
pre-date the last glacial cycle, and are therefore beyond the scope of this study.
The two younger glaciations, however, occurred both during the last glacial
cycle. First, the older of the two, the Gladstone Glaciation, reached its maximum extent between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during Marine Oxygen Isotope Stage (MIS) 4, a period of relatively large global ice volume. However, evidence from the Gladstone Glaciation has not been
documented along the southern ice sheet margin, making it difficult to
seize its spacial extent. The younger of the two last glacial cycle glaciations, the McConnell Glaciation, is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which appears to correspond to the last global ice volume maximum (MIS~2). In contrast to
the Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with glacial advance of the southern ice sheet margin during the Fraser
glaciation, with a maximum extent attained between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}.
Within errors inherent to the cosmogenic nuclide exposure dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet. Apart from geological evidence of the final demise of the ice sheet, most field evidence relate to the LGM maximum extent.

Ice has covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The ice sheet originated from the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}. It is generally thought that the LGM configuration of the Cordilleran ice sheet was short-lived
    \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a hypothesis which can be studied through numerical modelling.

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of coalescent piedmont
lobes, each of which formed at the outlet of the main valleys. The Puget Lobe
\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest Lobes. It covered the Puget Lowland, a region that comprises
the cities of Vancouver and Seattle. Further east, beyond the Cascade Range, the
Okanagan and Purcell lobes formed the southern Cordilleran ice sheet margin. The latter once dammed the Clark Fork River, inhibiting runoff and forming
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet margin retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The \emph{western margin} of the Cordilleran ice sheet covered part of the Pacific continental
shelf. An independent
ice cap covering Vancouver Island diverted flow from the main body of ice to drain into the Strait of Georgia between Vancouver Island and the Olympic Peninsula to form the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during maximum of glaciation. Submarine troughs traversing several portions of the continental shelf indicate the erosional action of
additional ice streams, yet it is not clear whether they have formed during the
last glacial cycle or during earlier glaciations \mref.

Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice sheet are the least understood.
Although it
is known that the Cordilleran ice sheet collided with the Laurentide ice sheet at times during the last glacial cycle,
the nature of the interaction between the two ice sheets and the precise
location of this junction remains unclear \citep{Gowan.2013}, while limited information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, I, therefore, neglect buttressing effect of the
Laurentide ice sheet.

The retreat of the ice sheet during the last deglaciation was likely rapid. Most of the evidence for the style and pace of deglaciation comes
from the Fraser Plateau. Studies here have indicated that some of the highest relief
emerged from the ice prior to the surrounding lowlands, possibly resulting in thin,
stagnant ice remnants covering parts of the plateau
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Less is known about the deglaciation of the northern sector of the Cordilleran ice sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of \emph{deglaciation}, with active retreat of the ice margin
towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.

\todo{To make this more complete I suggest two additions: 1. a few sentences regarding the timing of deglaciation (to complement your introduction where you have timing of glaciation) - look in my 2010 and Martin's papers for information. 2. A few more references older than Margold et al. 2013. You might check what Ward has written fordeglaciation in the Yukon for example? Didn't Dyke write about the deglaciation of the Liard Lowland?}

@@

The simulations presented in this thesis employ the Parallel Ice Sheet Model (PISM),
an open-source, finite-difference, shallow ice sheet model
\citep{PISM-authors.2014}. The model requires bedrock topography, sea level,
geothermal heat flux and climate forcing as inputs, and computes the thermal
and dynamic state of the ice sheet, and the associated thermal and
deformational responses of an idealised lithosphere. However, the atmosphere and ocean
are not part of the model. Instead, PISM includes boundary models
that provide simple parametrisations of their interaction with the ice sheet across interfaces that are assumed to be well defined (\cref{fig:model-interfaces}).

This section does not attempt to provide an extensive description of PISM,
which can be found in an on-line documentation \citep{PISM-authors.2014}.
Instead, it aims to highlight modelling choices made in the context of this
thesis, and presents an overview of the physics involved in these choices.

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and
conservation of mass therefore implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
capped at a threshold of 0.01, above which no measurements are available
\citep{Lliboutry.Duval.1985}.

\todo{I am trusting PISM's manual on that. Check other papers by Duval, or
      the book by Greve (library). Explain params (Arjen + Irina) or refer to
      table, maybe.}

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to
account for thermodynamic effects associated with internal phase changes in the
glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$) formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, $T_c$, the enthalpy can be
expressed as $H=c\,T$, where, $c$, denotes the specific heat capacity of the ice, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$, where $k$ is its thermal conductivity of ice. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,.
\end{equation}
Parameter values for the ice flow model used in this thesis are
given in \cref{tab:iceflowparams}.

\todo{Explain $H=c\,T$. Enthalpy here is equivalent to internal energy.}

\Cref{eqn:incompressibility,eqn:stressbalance,eqn:glenslaw,eqn:enthalpy}
constitute the thermodynamic basis
of the model. However, their resolution in full form implies a computational
demand too high for multi-millennial, continental-scale applications, such as
the numerical modelling of the Cordilleran ice sheet through the 120\,ka last glacial
cycle, thus requiring shallow approximations.

@@

PISM employs two well-documented
approximations of ice flow equations: the shallow ice approximation (SIA)
and the the shallow shelf approximation (SSA). Although both approximations can
be derived from a rigorous scaling approach
    \citep{Morland.Johnson.1980, Hutter.1983,
           Morland.1987, Weis.etal.1999},
I won't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

Neglecting atmospheric pressure, \Cref{eqn:siaz} yields \todo{detail}
\begin{equation}
    p = \rho g (s-z) \,,
\end{equation}
which can be re-introduced into \cref{eqn:siax,eqn:siay}, thus yielding the
expression of the horizontal shear stresses, sometimes referred to as driving
stresses,
\begin{subequations}
\begin{align}
    \tau_{xz} &= -\rho g (s-z) \pdv{s}{x} \,, \\
    \tau_{yz} &= -\rho g (s-z) \pdv{s}{y} \,.
\end{align}
\end{subequations}

@@

\todo{explain parameters}

Drainage of water at the base of the ice sheet, although implemented in PISM,
is not included in simulations presented here for the sake of simplicity.
